Learning how to act: making good decisions with machine learning

### 2017-05-18 thesis/thesis_map.tex
updates to chapter 2

\documentclass[11pt,a4paper,oneside]{book}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,positioning} 
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[numbers]{natbib}
\usepackage{wrapfig}

@@

%\tikzset{
%    %Define standard arrow tip
%    >=stealth',
%    %Define style for boxes
%    punkt/.style={
%           circle,
%           rounded corners,
%           draw=black, thick,
%           text width=1em,
%           minimum height=1em,
%           text centered},
%    punktrect/.style={
%    rectangle, 
%    rounded corners, 
%    % fill=black!10,
%    draw=black, thick,
%    minimum height=1em, 
%    text centered},
%    % Define arrow style
%    pil/.style={
%           o->,
%           thick,
%           shorten <=2pt,
%           shorten >=2pt,}
%}

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

@@

%\theoremstyle{plain}

\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

%\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\let\temp\epsilon
\let\epsilon\varepsilon

@@

There is a connection here to the degree to which the model is invariant. A model that is invariant to more things is more explainitory. We could look at levels of hierachy. There is causal graphical models, which are invarient to interventions on variables. Can I tie this to a physical system and give more indepth examples of the nested tier of hieriachies?

I do not see the distinction between explaination and (causal) prediction. Explaination is all about the ability to compress and to generalize. The more a model can do this, the more we view as providing an understanding of the why.

@@

Relationship to generalization. Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time. The assumption is there are less places for change to come in.

@@

Generalizing the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parashutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalize). (and given sufficient generalizability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)

@@

Only from randomized experiment ... if people can do it without recource to constant randomized experiment then an algorithm exits.

The challenges and limitations of infering causal effects from observation data are numerous. However, there are problems we need to solve. The questions is are these problems so serious as to warrent discregarding all non-experimental data. I think answer is a clear no. Need ongoing validation, with assumptions. We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).

@@

Are causal/interpretable models more reliable or more likely to generalize well?

@@

Specifically, people can
\begin{itemize}
\item rule out many possible features as highly unlikely to be relevent to a problem
\end{itemize}

@@

This is tightly related to generalizability. If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results. We don't include , the color of the doctors shirt, the size of the room, ...

@@

\chapter*{Introduction (3500 words)}

@@

\paragraph*{This thesis contributes to knoweldge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.

\paragraph*{My key research questions are:} 
\begin{itemize}
\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
\item To summarize the key strands of causal inference research from economics and social sciences for the machine learning community
\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
\end{itemize}

\section*{Motivation}
Many of the most important questions in science and in our personal lives are about the outcomes of doing something. Will asking people to pay upfront at the doctors reduce long term health expenditure? If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.

@@

The rise of big data sets and powerful computers has seen an explosion in the application of machine learning. From healthcare, to entertainment and self driving cars; machine learning algorithms will transform many industries. It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge datasets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artifcial intelligence exceeds our own and then grows exponentially.

However, despite the huge advances in specific areas of machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings. Getting them to generalize to even slightly different problems or datasets remains very challenging. Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it. The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on. As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.

\section*{What is causality and why do we care? (2000 words)}

@@

\paragraph*{There are two reasons why correlation is not causation}. The first is related to variance and over-fitting. Obversations are noisy. With a finite dataset with enough variables we will be able to find some that are completely unrelated but correlate purely by chance. \ref{figure - search for random correlation}. The second arises from bias, typically introced by an un-observed confounding variable. In this case, variables are correlated not by chance. We would expect the relationship to hold if we sampled more data. However, they are not causally related in that intervening to set one would not likely effect the other. EXAMPLE WITH FIGURE.

\subsection*{Defining causality}
\begin{itemize}
\item widely debated in science and philosophy (REFERENCES)
\item what is explaination?
\item any model that aims to predict the outcome of an action or intervention in a system
\item I do not see the distinction between explaination and (causal) prediction. Explaination is all about the ability to compress and to generalize. The more a model can do this, the more we view as providing an understanding of the why. 
\item mediation?
\end{itemize}

\subsection*{Identifying when we have a causal problem}

\subsubsection*{Examples of typical machine learning probems. Are they causal?}

Consider the following problems, which span a wide range of the types of questions machine learning is currently being applied to. Which of them require casual inference? How can we identify characterists of a prolem that make causal modeling important?

\begin{itemize}
\item Speech recognition (for systems like Siri or Google)
\item Machine translation 
\item Image classification
\item Forecasting the weather
\item Playing Go 
\item Identiying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pnumonia.
\item Predicting who will re-offend on release from prison 
\item Predicting which customers will cease to be your customers
\item Demand prediction for inventory control
\item Predicting who will click on an ad
\item Financial trading
\item Recommending movies
\item Online search
\item Self driving cars
\item Pricing insurance
\end{itemize}

@@

Consider speech recognition. You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words. Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a dvd on elocution). A similar arguament could be made for many applications of machine translation and image classification.

In image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see. We can be fairly confident that the process will not change. If we develop a discriminative model that is hightly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).

What about forecasting the weather? If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decicion will not effect if it actually rains. However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system. For this we need a (causal) model that allows us to predict the outcome under various different interventions.

Identifing spam and automated essay marking systems are similar. The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label. Spammers and students will modify their writing in order to optimise thier results. The standard machine learning approach can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided. (What would the nature of the features have to be such that change did not occur? - they would have to be causes of the label).

What about predicting the risk of death in patients with pnumonia? Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective this seems like a straighforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or ommited) in the model, the relationship between those features and the outcome may change if we start using the model to decide whome to admit. (xxx et al) found exactly this effect. Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pnemonia. They realised this was because doctors were treating such patints very agressively, thus acutally lowering their risk. There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addmition and treatment protocols. However, using it to decide on what basis to admit people could kill. The key is understanding exactly what question you are asking. In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.

@@

Demand prediction seems relatively straightforward. These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product. It seems unlikely that using the model to ensure stock is available will itself change demand. However there is a potential data sensoring issue. If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero. Changing availability does then change demand.

Playing Go (and other games) is another case with some subleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win. ... this is interesting maybe come back to it ... an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ... One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken). When can we learn from the actions others have taken? When there is no confounding. And does this hold with Go? Probably because the board state encapsulates everything that should determine what move is played. Learning directly from actions and trying to generalize (can in some instances reduce the problem to standard ML)

Having considered these examples we can now identify some general aspects of the problem that determine wheter or not we require a causal model.

@@

 Covariate shift clearly comes in here. Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occuring and that results will be transferable. The mechanism is known but the function may be complex. Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models? Yes, if the way in which features get their values changes, then causal models will be invarient to that in a way that non-causal ones are not.

\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and priortise those for modelling changes. 
\item Any decision we take does not significanly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
\item Ethics - ... People's viewpoint on if its ok...
\end{itemize}

I hope these examples gave you a feel for the richness and subtlties of causal inference. We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.

\section*{Approaches to causality (1000 words)}

@@

We will call datasets where we do not have control over the decision making process that generated the data observational. Versus experimental datasets, where we do have control (experimental datasets do not always have to be randomized -although that is a powerful approach to ensure we have control. There is a space in the middle where we have partially controlled the process by which agent select actions. Randomized data with imperfect compliance would be an example.

An agent (capable of intervening in the system) chooses an action from those availble The agent making the decision included in the model. This is

@@

There are two key approaches to causal problems. The first is to learn the outcome of actions by directly intervening in the system and seeing what happens. We then get feedback on how good those actions were. This is the approach taken in reinforcement learning. ADVANTAGES AND DISADVANTAGES OF THIS APPROACH. The second broand approach is causal inference. Here

@@

\paragraph*{Both generalize from randomized experiment} Reinforcment learning to sequential decisions, causal inference to non-experimental conditions

@@

\paragraph*{Limitations of experimets} What are the issues with standard randomized experiments?

\chapter*{Causal models (3000 words)}

Causal inference aims to infer the outcome of an intervention in a system from data obtained by observing (but not intervening in) the system. To do this we need terminology to describe actions and how we anticipate the system should respond to them. Three key approaches have emerged; counterfactuals, structural equation models and causal bayesian networks. In this chapter we describe these approaches. We will examine what problems they allow us to solve, what assumptions they rely on and how they differ. We will also use them to describe the following very simple example. The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.

Suppose we have developed a new drug for some illness and wish to determine how effective it is. We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group. The people in the treatment group get the drug, everyone else gets a placebo pill. The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo. We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome.

\subsection*{Causal bayesian networks}

Causal Bayesian Networks are an extension of Bayesian networks. A bayesian network is a graphical way of representing how a distriution factorises. Any joint probability distribution can be factorized into a product of conditional probabilities. There are multiple valid factorizations, corresponding to permutations of variable ordering.

\begin{equation}
P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
\end{equation}

We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution (figure~\ref{fig:bayesnet}). If there are conditional independencies between variables the factorization simplifies, which is reflected by missing edges in the corresponding network.

\begin{figure}[h]
\centering
\caption{General Bayesian network over 3 variables}
\label{fig:bayesnet}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

@@

There are some complex philosopical objections to counter factuals arising from the way they describe alternate universes that were never realized and are not empirically testable \ref{XXXXX}. This can result in some concrete problems. Consider the following example based on  \ref{Dawid2000}.

@@

- does it actually make sense to make a decison on the basis of this counterfactual that is a function of something we could never observe. I

- part of the problem may be due to the deteminisitic way we have phrased individual treatment effects.

Can these issues be resolved by considering personalized causal effects as random variables and avoiding counterfactuals?

>>>>

\documentclass[11pt,a4paper,oneside]{book}
\usepackage[table]{xcolor}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{algorithm}% http://ctan.org/pkg/algorithm
\usepackage[noend]{algorithmic}
\usepackage{parskip}
\usepackage{tikz}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{epigraph}
\usetikzlibrary{arrows,fit,positioning}
\usepackage{subcaption}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[numbers]{natbib}
\usepackage{wrapfig}

% this is included to get tables with colored rows to not draw over hlines and vlines
\usepackage{etoolbox}

\makeatletter
\patchcmd{\@classz}
  {\CT@row@color}
  {\oldCT@column@color}
  {}
  {}
\patchcmd{\@classz}
  {\CT@column@color}
  {\CT@row@color}
  {}
  {}
\patchcmd{\@classz}
  {\oldCT@column@color}
  {\CT@column@color}
  {}
  {}
\makeatother

\usepackage{amsthm}

@@

\tikzset{
    %Define standard arrow tip
    >=stealth',
    %Define style for boxes
    punkt/.style={
           circle,
           rounded corners,
           draw=black, thick,
           text width=1em,
           minimum height=1em,
           text centered},
    observed/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
     latent/.style={
           circle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
    target/.style={
           circle,
           rounded corners,
           draw=black, thick,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
    observedrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!10!white
           },
    latentrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick, dashed,
           minimum width=2.2em,
           minimum height=2.2em,
           font=\footnotesize,
           text centered
           },
     targetrect/.style={
           rectangle,
           rounded corners,
           draw=black, thick,
           minimum width=6em,
           minimum height=2em,
           font=\footnotesize,
           text centered,
           fill=black!20!white,
           },
     empty/.style={
           circle,
           rounded corners,
           minimum width=.5em,
           minimum height=.5em,
           font=\footnotesize,
           text centered,
           },
    % Define arrow style
    pil/.style={
           o->,
           thick,
           shorten <=2pt,
           shorten >=2pt,},
    sh/.style={ shade, shading=axis, left color=red, right color=green,
    shading angle=45 }   
}

@@

\theoremstyle{plain} 
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\let\temp\epsilon
\let\epsilon\varepsilon

@@

\begin{figure}[h]
\centering
\caption{Causal inference without a framework}
\label{fig:mutilatednet}
\begin{subfigure}[t]{0.3\textwidth}
\centering
\caption{original network, $G$}
\label{fig:mutilatedOriginal}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$U$};
\node[main node, below left=of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};

\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.5\textwidth}
\centering
\caption{network after an intervention $G^*$}
\label{fig:mutilatedAfter}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

 %nodes
\node[main node](1){$U$};
\node[main node, below left=of 1](2){$x$};
\node[main node, below right=of 1](3){$Y$};

\begin{figure}[h]
\caption{}
\label{fig:causal_inference_without_do}
\centering
\begin{tikzpicture}[->,shorten >=0pt,shorten <=0pt,node distance=2.5em,thick,node/.style={observedrect},lt/.style={latent}]
\node[node](2){Treatment};
\node[node, right=of 2](3){Outcome};
\node[lt, above=of 3](4){U};
\path[]
	(2) edge (3)
	(4) edge (3);

There is a connection here to the degree to which the model is invariant. A model that is invariant to more things is more explanatory. We could look at levels of hierarchy. There is causal graphical models, which are invariant to interventions on variables. Can I tie this to a physical system and give more in depth examples of the nested tier of hierarchies?

I do not see the distinction between explanation and (causal) prediction. Explanation is all about the ability to compress and to generalise. The more a model can do this, the more we view as providing an understanding of the why.

@@

Relationship to generalisation. Variables causally directly causally related to the outcome (either causes or effects) should be more stable predictors over time. The assumption is there are less places for change to come in.

@@

Generalising the results of one experiment to another (for example, dropping rocks to dropping people - with reference to the cross over trial for parachutes.) (This I think is the long term key to successful causal inference, learning from experimental data with representations that generalise). (and given sufficient generalisability, it may not matter if there is an underlying confounder - as this confounder is clearly not changing much)

@@

Only from randomised experiment ... if people can do it without recourse to constant randomised experiment then an algorithm exits.

The challenges and limitations of inferring causal effects from observation data are numerous. However, there are problems we need to solve. The questions is are these problems so serious as to warrant disregarding all non-experimental data. I think answer is a clear no. Need ongoing validation, with assumptions. We can't validate the assumptions using the observational data alone (otherwise we would not have been forced to make them).

@@

Are causal/interpretable models more reliable or more likely to generalise well?

@@

Specifically, people can
\begin{itemize}
\item rule out many possible features as highly unlikely to be relevant to a problem
\end{itemize}

@@

This is tightly related to generalisability. If we take a person undergoing a medical test, we might describe the situation by the year and location, the person's age, gender, heart rate, medical condition and test results. We don't include , the color of the doctors shirt, the size of the room, ...

@@

\chapter{Introduction}

@@

\paragraph*{This thesis contributes to knowledge by:} Introducing a framework connecting causal graphical models with multi-armed bandits as a first step towards a unified approach to estimating the effect of interventions.

\paragraph*{My key research questions are:} 
\begin{itemize}
\item To understand and the difference between prediction and causal inference in machine learning and clarify which problems require causal approaches.
\item To summarise the key strands of causal inference research from economics and social sciences for the machine learning community
\item To make connections between learning to act from observational versus experimental data. In particular, between causal graphical models and multi-armed bandits.
\end{itemize}

\section{Motivation}
Many of the most important questions in science and in our personal lives are about the outcomes of doing something. Will asking people to pay upfront at the doctors reduce long term health expenditure? If we developed a drug to suppress particular genes, could we cure MS and would delaying teen-aged pregnancies improve the outcome for their kids.

@@

The rise of big data sets and powerful computers has seen an explosion in the application of machine learning. From healthcare, to entertainment and self driving cars; machine learning algorithms will transform many industries. It has been suggested that the impressive ability of statistical machine learning to detect complex patterns in huge data sets heralds the end of theory (Reference) and that we may be only a short step from the Singularity, where artificial intelligence exceeds our own and then grows exponentially.

However, despite the huge advances in specific areas of machine learning (in particular deep learning), machine learning algorithms are effective only within very narrow problem settings. Getting them to generalise to even slightly different problems or data sets remains very challenging. Deciding how we should act or what policies we should implement requires us to be able to predict how a system will behave if we change it. The correlations detected by standard machine learning algorithms do not enable us to do this, no matter how many petabytes of data they are based on. As machine learning algorithms are incorporated into more and more of the decision making processes that shape the world we live it, it is critical to ensure that we understand the distinction between causality and prediction and that we develop techniques for learning how to act that are as effective as those we have for pattern recognition.

\section{What is causality and why do we care? (2000 words)}

The notion of causality has been widely debated in science and philosophy \cite{} but is still widely viewed as poorly defined \cite{}. This has led to a reluctance among applied researchers in many fields to refer to cauality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existance of an association depends on what other variables we adjust for (or include in a regression). Avoiding formalising causation, the real question of interest, leaves it up to the reader to determine via common sense if the association reported is the \emph{right one}.

We discuss more detailed definitions of causality in section c

The what-if type questions from the why. \cite{Gelman2010} \cite{}. Why do whites to better than blacks in school (in America). Suggests that reverse causal inference questions are more interesting and motivate most of social science.

I do not find this distinction useful. 
We can only change the future - history is useful only as far as it tells us something about the future.
Reverse causal questions can be reposed as forward ones, when making the translation a reverse causal query will be effectively asking about many possible interventions (rather than just one). Problems highlighted as intractible in the reverse causal sense are also intractible in the forward inference form, typically because concern situations for which we do not have a sufficient number of similar instances to allow statistical reasoning. For example, the war question posed in Gelman.

A distinction between forward causal iference, what happens if we do X and reverse casual inference

@@

\paragraph*{There are two reasons why correlation is not causation}. The first is related to variance and over-fitting. Observations are noisy. With a finite data set with enough variables we will be able to find some that are completely unrelated but correlate purely by chance. \ref{figure - search for random correlation}. The second arises from bias, typically introduced by an un-observed confounding variable. In this case, variables are correlated not by chance. We would expect the relationship to hold if we sampled more data. However, they are not causally related in that intervening to set one would not likely effect the other. EXAMPLE WITH FIGURE.

\subsection{Defining causality}
\begin{itemize}
\item widely debated in science and philosophy (REFERENCES)
\item what is explanation?
\item any model that aims to predict the outcome of an action or intervention in a system
\item I do not see the distinction between explanation and (causal) prediction. Explanation is all about the ability to compress and to generalise. The more a model can do this, the more we view as providing an understanding of the why. 
\item mediation?
\end{itemize}
\subsection{Overview of this thesis}
\subsection{Identifying when we have a causal problem}

\subsubsection{Examples of typical machine learning problems. Are they causal?}

Consider the following problems, which span a wide range of the types of questions machine learning is currently being applied to. Which of them require casual inference? How can we identify characteristics of a problem that make causal modelling important?

\begin{itemize}
\item Speech recognition (for systems like Siri or Google)
\item Machine translation 
\item Image classification
\item Forecasting the weather
\item Playing Go 
\item Identifying spam emails
\item Automated essay marking
\item Predicting the risk of death in patients with pneumonia.
\item Predicting who will re-offend on release from prison 
\item Predicting which customers will cease to be your customers
\item Demand prediction for inventory control
\item Predicting who will click on an ad
\item Financial trading
\item Recommending movies
\item Online search
\item Self driving cars
\item Pricing insurance
\end{itemize}

@@

Consider speech recognition. You say something, which causes to sound waves, which are converted to a digital signal which Siri maps to words. Whatever action Siri takes is unlikely to change the distribution of words you use, and even less likely to change the function that maps sound waves to text (unless she sends you a DVD on elocution). A similar argument could be made for many applications of machine translation and image classification.

In image recognition, we do not particularly care about building a strong model for exactly how the thing that was photographed translates to the image we see. We can be fairly confident that the process will not change. If we develop a discriminating model that is highly accurate at classifying cats from dogs, we do not care a lot about its internal workings (provided we have strong grounds to believe that the situations in which we will be using our model will match those under which it was trained).

What about forecasting the weather? If you are using a short term forecast to decide whether to pack an umbrella it's clear causality can be ignored - your decision will not effect if it actually rains. However, longer term climate forecasts might (theoretically) lead us to take action on emissions which would then change the weather system. For this we need a (causal) model that allows us to predict the outcome under various different interventions.

Identifying spam and automated essay marking systems are similar. The decision made by the algorithm is likely to change the relationship between the features used by the algorithm and the true label. Spammers and students will modify their writing in order to optimise their results. The standard machine learning approach can only work if the resulting change is sufficiently gradual and fresh ground truth (probably human labelled) training data is provided. (What would the nature of the features have to be such that change did not occur? - they would have to be causes of the label).

What about predicting the risk of death in patients with pneumonia? Suppose we wish to use the model to decide who should be treated in hospital and who can be sent home with antibiotics. If we assume that in hospital treatment is more effective this seems like a straightforward prediction problem. It is not. Depending on how the decision to admit was previously made and what features are included (or omitted) in the model, the relationship between those features and the outcome may change if we start using the model to decide whom to admit. (xxx et al) found exactly this effect. Their model learnt that (among other things) people suffering asthma were \emph{less} likely to die from pneumonia. They realised this was because doctors were treating such patients very aggressively, thus actually lowering their risk. There is no problem with this model if you want to predict who would be likely to die whilst maintaining the original addmition and treatment protocols. However, using it to decide on what basis to admit people could kill. The key is understanding exactly what question you are asking. In this case we are care about what happens to patients with characteristics X if we treat them according to decision rule Z.

@@

Demand prediction seems relatively straightforward. These models use features such as location, pricing, marketing, time of year, weather, etc to forecast the demand for a product. It seems unlikely that using the model to ensure stock is available will itself change demand. However there is a potential data censoring issue. If demand is modelled by the number of sales, then if a product is out of stock demand will appear to be zero. Changing availability does then change demand.

Playing Go (and other games) is another case with some subtleties. At every turn, the AI agent has a number of actions available. The board state following each action is deterministic and given by the rules of the game. The agent can apply supervised machine learning based on millions of previous games to estimate the probability that each of these board states will lead to a win. ... this is interesting maybe come back to it ... an alternate approach would be to try to forecast the probability of a win given each action given the current board state as context ... One approach to causal inference is indeed to learn about actions from taking actions (or observing the actions that other have taken). When can we learn from the actions others have taken? When there is no confounding. And does this hold with Go? Probably because the board state encapsulates everything that should determine what move is played. Learning directly from actions and trying to generalise (can in some instances reduce the problem to standard ML)

Having considered these examples we can now identify some general aspects of the problem that determine whether or not we require a causal model.

@@

 Covariate shift clearly comes in here. Because there are areas where mechanisms are understood it is relatively easy to argue that covariate shift is not occurring and that results will be transferable. The mechanism is known but the function may be complex. Can we write down something that causal models are invariant to in terms of shift that is not the case for non-causal models? Yes, if the way in which features get their values changes, then causal models will be invariant to that in a way that non-causal ones are not.

\item To decide between actions we only need to rank them (not estimate their actual effect). 
\item The predicted outcome in the absence of an intervention provides a single point. We can use this to find which problems are most serious if left alone - and prioritise those for modelling changes. 
\item Any decision we take does not significantly impact the system from which the data was drawn to make it (for repeat decision making)
\item Does acting on the result of the prediction change the predictive distribution p(y|x)? Ie change people's behaviour.
\item Ethics - ... People's viewpoint on if its OK...
\end{itemize}

I hope these examples gave you a feel for the richness and subtitles of causal inference. We will return to some of them in more detail once we have established some more concrete language and tools to approach them with.

\section{Approaches to causality (1000 words)}

@@

We will call data sets where we do not have control over the decision making process that generated the data observational. Versus experimental data sets, where we do have control (experimental data sets do not always have to be randomised -although that is a powerful approach to ensure we have control. There is a space in the middle where we have partially controlled the process by which agent select actions. Randomised data with imperfect compliance would be an example.

An agent (capable of intervening in the system) chooses an action from those available The agent making the decision included in the model. This is

@@

There are two key approaches to causal problems. The first is to learn the outcome of actions by directly intervening in the system and seeing what happens. We then get feedback on how good those actions were. This is the approach taken in reinforcement learning. ADVANTAGES AND DISADVANTAGES OF THIS APPROACH. The second broad approach is causal inference. Here

\todo{insert figure showing data generating process and observed data defining ML}\ref{all of statistics}

@@

\paragraph*{Both generalise from randomised experiment} Reinforcement learning to sequential decisions, causal inference to non-experimental conditions

@@

\paragraph*{Limitations of experiments} What are the issues with standard randomised experiments?

\chapter{Causal models}
\label{chap:causal_models}

Causal inference aims to infer the outcome of an intervention in some system from data obtained by observing (but not intervening in) it. To do this we need terminology to describe actions and how we anticipate the system should respond to them. Three key approaches have emerged; counterfactuals, structural equation models and causal bayesian networks. In this chapter we describe these approaches. We will examine what problems they allow us to solve, what assumptions they rely on and how they differ. We will also use them to describe the following simplified examples. The aim is to demonstrate the notations and formalisms we will need to tackle more interesting problems later on.

\vspace*{.3cm}
\begin{example}
\label{exm:ranomized_experiment}
Suppose we have developed a new drug for some illness and wish to determine how effective it is. We take a large group of patients and randomly assign half of them to a treatment group and the other half to a control group. The people in the treatment group get the drug, everyone else gets a placebo pill. The question we want to answer is does giving people the active drug improve their changes of recovery relative to giving them the placebo. We will use the variable $X$ (1 = drug, 0 = placebo) to represent the treatment each person receives and $Y$ (1 = recover, 0 = not recover) to describe the outcome. 
\end{example}

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we wish to estimate what would happen to high school graduation rates if we made pre-school available and compulsory for all 4 year olds. We have a large cross-secional dataset on a group of 20 year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.  
\end{example}

\section{Causal bayesian networks}

Causal Bayesian networks are an extension of Bayesian networks. A Bayesian network is a graphical way of representing how a distribution factorises. Any joint probability distribution can be factorised into a product of conditional probabilities. There are multiple valid factorisations, corresponding to permutations of variable ordering.

\eqn{
\label{eqn:cbn:joint_dist}
P(X_{1},X_{2},X_{3},...)=P(X_{1})P(X_{2}|X_{1})P(X_{3}|X_{1},X_{2})...
}

We can represent this graphically by drawing a network with a node for each variable and adding links from the variables on the right hand side to the variable on the left for each conditional probability distribution, see figure \ref{fig:bayesnet}. If the factorisation simplifies due to conditional independencies between variables, this is reflected by missing edges in the corresponding network. There are multiple valid Bayesian network representations for any probability distribution over more than one variable, see figure \ref{fig:bayesnet2} for an example.

\begin{figure}[h]
\centering
\caption{A general Bayesian network for the joint distribution over three variables. This network does not encode any conditional independencies between its variables and can thus represent any distribution over three variables.}
\label{fig:bayesnet}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,main node/.style={punkt}]

@@

There are some complex philosophical objections to counter factuals arising from the way they describe alternate universes that were never realised and are not empirically testable \ref{XXXXX}. This can result in some concrete problems. Consider the following example based on  \ref{Dawid2000}.

@@

- does it actually make sense to make a decision on the basis of this counterfactual that is a function of something we could never observe. I

- part of the problem may be due to the deterministic way we have phrased individual treatment effects.

Can these issues be resolved by considering personalised causal effects as random variables and avoiding counterfactuals?
